Attrition in Massive Open Online Courses

### 2015-01-12 paper_v2.tex
Updating paper after first review

We fit logistic mixed-effects (hierarchical) models to a number of binary persistence and performance metrics to estimate individual differences in prior learner characteristics, including demographic and geographic features. Correlations between predictors were low enough to allow for simple interpretation of coefficients. Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{The required data for 3 of the 17 MOOCs could not be obtained in time for submission due to technical issues.} Figure \ref{fig:s1coefs} illustrates regression coefficients with 95\% confidence intervals from each model. The coefficients represent changes from a baseline (intercept) population: male learners in North America, with an average age, who had not earned a bachelor's degree and had no prior experience with the topic or taken other MOOCs, not intending to complete the course, and prepared to spend only up to two hours per week on the course.

A number of significant and substantial individual differences emerged, some indicative of potential inequalities: Women were 10 to 15\% less likely than men to persist with lectures and assessments. Women are also less likely to score a grade above the 60$^\text{th}$ percentile than men. Learners in Africa, Asia, and South America were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America. In contrast, learners in Europe exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America. While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America.

Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance. More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to perform highly on them. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance. Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.

@@

\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop]{Coefficients of predictors (binary, except for normalized age) from logistic mixed-effects model predicting persistence and performance milestones (prop. videos watched, prop. assessments completed, and grade percentile). The baseline group were male learners in North America of average age who had not earned a bachelor's degree, without prior experience in the topic or other MOOCs, not intending to complete the course, and prepared to spend under two hours per week on it. An estimate of -20 indicates a 20\% lower probability of achieving a milestone.\label{fig:s1coefs}}
\end{figure*}

@@

Attrition can be a sign of success if a learner has achieved their personal goal. We therefore investigated to what extent learners were satisfied with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$). To account for differences across courses, we compared satisfaction scores using mixed-effects models.

Learners who reported that they stopped participating were significantly 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=12, p<.001$), it was a relatively weak effect. Those who watched all videos were only around 3\% more satisfied than those who watched none, based on a linear fit.

@@

Research question $RQ3a$ about learners' reasons for disengaging from a MOOC is addressed by examining the proportion of learners who selected reasons for attrition from a pre-defined list.\footnote{This researcher-defined list may not have been collectively exhaustive, which is addressed in Study 2.} The distribution of proportions across 17 MOOCs is illustrated in Figure \ref{fig:s1reason} with boxplots. Two reasons related to having too little time stand out because they were reported by about half of the respondents as influencing their decision to stop participating: the reasons were issues keeping up with deadlines, and the course demanding too much time. Other reasons for attrition were less prevalent, though still reported by around 10\% to 25\% of respondents. Notably, just under 20\% of respondents in a typical course (the median proportion) stopped participating because they had learned what they wanted to learn.

To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em  difficulty} (too advanced, exam too difficult); fourth, {\em format and content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable). Only the technical difficulties item failed to load onto any factor.

We expected reasons for disengaging from the course to vary by learner characteristics ($RQ3b$), such as learners' motivations for enrolling in the course. A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations. Taking the course with friends or colleagues was correlated with disengaging because the course is not challenging enough, the material is too advanced of the learner's background, and the credential is not valuable enough to the learner ($r=.24, .20, .16$, respectively). Not surprisingly, learners who enrolled to experience an online course tended to stop participating because they were only exploring the course ($r=.17$). Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).

@@

A number of striking findings emerged from the preceeding analyses. We found strong evidence for gender inequalities in attrition and, to some extent, performance. Women were less likely to persist, even when adding a number of other demographic, geographic, and preference indicators to the model, which highlights the robustness of the relationship. The MOOCs in our dataset were on diverse topics, including some on topics in the humanities and business with high rates of participation from women. Besides gender differences, attrition was substantially higher for learners from Africa, Asia, and South America, compared to North America, Oceania, and Europe, in terms of lectures watched. We also identified a large achievement gap in terms of grades between learners from these continents. The current generation of MOOCs is seemingly not immune to inequalities that have pervaded traditional educational settings. It remains unclear, however, to what extent these gaps result from differences in Internet access and language barriers, or potentially, from experiences of psychological threat to learners' identity or sense of social belonging \cite{walton2007question}.

Levels of self-reported satisfaction were generally high and did not vary much by level of persistence, although learners who reportedly disengaged from the course tended to be less satisfied than those who reportedly persisted until the end. Learners' reasons for disengaging from the course clustered into five key factors: general time issues; problems due to starting the course late; course difficulty; issues with the course format or content, and a mismatch or fulfillment of expectations. General time issues were overall the most pronounced.

@@

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. To this end, we developed a predictive model to identify disengaging learners and asked them to complete a survey. One goal of this study was to further investigate research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced. Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time. A variety of factors can influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments, are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. For instance, learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV). We constructed the following hypotheses about the presence of these two factors for learners who report having not enough time:

@@

Another goal of Study 2 was to collect data to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners. To this end, we included a self-report measure for each construct in the survey that was sent to disengaging learners.

@@

The particular MOOC under observation was an undergraduate level course on an advanced topic in computer science. It was offered in 2014 through Coursera. There were  20,048 enrolled learners; 10,510 watched at least one video, and 5,912 attempted at least one assignment. The following subsections describe how disengaging learners were identified, which measures were included in the feedback survey, and how open responses were iteratively coded.

@@

The training set was composed of learners' interaction data (predictors) and disengagement states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see omitted for blind review%\cite{halawa2014dropout}
, for additional information on the prediction model).

To control the length of the prediction lag, the model was only provided with learners' activity data until they were absent for 14 consecutive days. In the training phase, if the learner disengaged but the model did not detect it within 14 days of their last activity, the case was counted as a false negative. This construction results in a model with features that all fall within the defined prediction lag period. The model was fit by logistic regression with two-fold learner-level and course-level cross-validation. Table \ref{tab:predAuc} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

\begin{table}[h!]
\caption{Performance of the attrition prediction model}
\label{tab:predAuc}
\small
\center
\begin{tabular}{lccc}
\toprule
 & $train$ & $test_1$ & $test_2$ \\
\midrule
AUC &  0.931 & 0.929 & 0.922  \\
\bottomrule
\end{tabular}
\end{table}

Two of the predictors in the fitted model are highly influential. The first is the number of days since the learner has last been active ($x_{1}$). The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more. However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed ($x_{2}$) prior to the absence; this is the second most influential feature in the model. Equation 1 is an approximation of the model with only these two predictors. The fact that almost identical results were obtained for $test_1$ and $test_2$ suggest good transferability of the model. A learner was flagged if their dropout probability exceeded a half:

\begin{equation}
Pr(\text{dropout}) = \frac{1}{1+e^{-0.431x_1 + 6.388x_2 + 1.626}} > 0.5
\end{equation}

Out of the 20,048 enrolled learners in the course, only those 10,510 were considered who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out. The survey was sent out to 6,053 learners (excluding 21 learners with invalid email addresses or duplicate accounts).

@@

Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked. These items served as measures of personal success. They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of. This question was deliberately asked prior to any survey questions that could suggest particular reasons for disengagement.

Two research assistants independently developed codebooks for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations. The codebook was updated in each iteration to reliably fit the open response data. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification. In the second iteration, the remaining responses were coded in the same way. In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see [URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.

% Learners also rated the extent to which their progress was hindered by a number of obstacles. 
Learners also rated their sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

@@

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. exhibit lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$).

Reasons for learner attrition were measured using iteratively coded open responses ($n=399$). Table \ref{tab:s2reas} provides each coded challenge and percentage of open responses that indicated it. A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the main challenge.

\begin{table}[h!]
\caption{Iteratively-Coded Reasons for Attrition}
\label{tab:s2reas}
\small
\center
\begin{tabular}{lr}
\toprule
What challenges have you experienced while taking this course?  & \% \\
\midrule
Not enough time for the course & 84 \\
The course format or teaching style is not a good fit & 18 \\
The level of the course is too advanced & 10 \\
The course is not interesting or valuable enough & 7 \\
Additional course materials needed to learn the topic & 4 \\
Technical difficulties & 2 \\
Language barrier & 1 \\
\bottomrule
\end{tabular}
\end{table}

To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time. It revealed that no particular commitment was specified in about half of them (53\%). This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses $H4a$ and $H4b$: that is, while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.

Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners. Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress. The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction. Hence, the results fully support hypothesis $H1$, and partially support $H2$ and $H3$.

@@

Study 2 focused on reasons for attrition and psychological factors of learners who were likely to drop out of the course. Through iterative coding, we developed a set of challenges that were faced by a subset of learners in a particular MOOC (cf. Table \ref{tab:s2reas}). The large majority of learners who indicate that they had not enough time was found to be separable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$). MOOCs are an environment that requires learners to be highly self-regulated to be successful. The present findings emphasize the critical role of their capacity for volitional control, a factor that warrants further investigation \cite{corno2001volitional}. Very few learners explicitly indicated issues with volitional control in their open responses, potentially because it is not socially desirable to admit. The absence of specific reasons for their shortage of time was found indicative of issues with volitional control ($H4b$). This could be a useful proxy for volitional issues in future research.

We investigated the extent to which personal success in MOOCs is associated with three established psychological factors: goal striving, social belonging, and growth mindset. Prior evidence found the absence of each factor to cause lower persistence and performance in more traditional achievement-oriented environments. The present findings highlight the critical role of these factors in MOOCs based on observational evidence. Learners with higher levels of goal striving, social belonging, and growth mindset also reported higher levels of personal success ($H1,H2,H3$). The subpopulation of learners in this study was a small and intentionally skewed subset. The observed differences are expected to be larger in the full population of learners, compared to this homogenous sample of learners who are unlikely to persist. Goal striving was most strongly associated with both measures of personal success. Success in terms of relative progress was associated with a stronger growth mindset, but not significantly with social belonging. In contrast, success in terms of satisfaction with progress was associated with stronger social belonging, but not significantly with growth mindset. Future work should examine if this pattern reflects a difference in the underlying process by which these psychological factors affect learner behavior and perception.

The methodology used in this study can be extended to deliver timely and targeted interventions. The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course. The majority of them were unsuccessful in their own terms. All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners. Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}). Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their identity or social belonging.

>>>>

We fit a logistic mixed-effects (hierarchical) model to predict multiple binary persistence and performance metrics (milestones) to estimate individual differences in learner characteristics, including demographic and geographic features. Correlations between predictors were low ($|r|<0.16$), except for age and education ($r=0.27$), which were each significant predictors if the other was excluded. The coefficient on education, for instance, reflects the association between having a bachelor's degree and achieving a milestone, partialing out all other predictors, including age. Fourteen MOOCs in the dataset were modeled as a random effect, in order to generalize to the population of similar MOOCs.\footnote{Data for 3 of the 17 MOOCs was unavailable due to anonymization.}

Figure \ref{fig:s1coefs} presents transformed regression coefficients with 95\% confidence intervals from each model. The estimates represent changes in the likelihood of achieving a milestone relative to an arbitrary baseline population. The regression intercept corresponds to the following baseline group that derives from the coding of regressors: average aged male learners in North America without a bachelor's degree, without prior experience with the topic or other MOOCs, prepared to spend up to 2 hours/week on the course, and not intending to complete it. In this sample of survey respondents, the baseline probability for each milestone was high: videos watched (27\% above 30\%, 16\% above 50\%, 8\% above 80\%) assessments attempted (78\% above 30\%, 55\% above 50\%, 33\% above 80\%), and grade (42\% above the 60$^\text{th}$, 31\% above the 80$^\text{th}$ percentile).

A number of significant and substantial individual differences emerged. Women were 10 to 15\% less likely than men to persist with lectures and assessments. Women, who constituted 41\% of learners in the sample, were also less likely to score a grade above the 60$^\text{th}$ percentile than men. Learners in Africa ($n = 2,097$), Asia ($n = 8,199$), and South America ($n = 2,574$) were 20 to 50\% less likely to persist in video watching and tended to achieve lower grades than learners in North America ($n = 18,353$). In contrast, learners in Europe ($n = 11,992$) exhibited 20 to 50\% higher persistence with lectures and assessemnts than learners in North America. While learners in Asia exhibited lower persistence with videos, they were more likely to attempt more assessments than learners from North America. Learners in Oceania ($n = 1,166$) performed similarly to those in North America and Europe.

Besides gender and geography, the older the learner, the more likely they were to persist in lectures and assessments, but the lower their relative performance. More educated learners, those with a bachelor's or higher degree, were somewhat more likely to start assessments, compared to those with less education, but were less likely to score highly on them. Not surprisingly, prior experience, intending to learn about all topics in the course, and being prepared to spend more hours in the course were strong predictors of persistance. Learners who took prior MOOCs, however, were less likely to persist, yet achieved higher grades.

@@

\includegraphics[width=\maxwidth]{figure/s1coefs} \caption[Estimates from logistic mixed-effects model predicting persistence and performance milestones]{Estimates from logistic mixed-effects model predicting persistence and performance milestones. An estimate of -10 indicates a 10\% lower chance of achieving a milestone.\label{fig:s1coefs}}
\end{figure*}

@@

Attrition can be a sign of success if a learner has achieved their personal goal. We investigated how satisfied learners were with their achievements for those who reported to have stopped participating ($RQ2a$) and those who participanted to different degrees ($RQ2b$). Satisfaction ratings were compared with a mixed-effects regression to model differences across courses and generalize to a larger population of MOOCs. Learners who reported that they stopped participating were 9\% less satisfied than those who reported not to have stopped ($t_{15140}=18, p<.001$). Moreover, we examined if persistent learners, those who watched more lectures, were also more satisfied. While the the proportion of videos significantly predicted higher satisfaction ($t_{15700}=14, p<.001$), the effect was relatively weak. Those who watched over 75\% of videos were only 5\% more satisfied than those who watched under 25\% ($M_{<\text{25\%}}=4.97, M_{>\text{75\%}}=5.22$).

@@

% Say how many actually reported N=?
Learners' reasons for disengaging from a MOOC were examined using survey responses ($RQ3a$). Across 17 MOOCs, learners selected reasons for disengaging from a pre-defined list. This list may not be collectively exhaustive, which is addressed in Study 2. The distribution of chosen reasons is illustrated in Figure \ref{fig:s1reason}. Around half of the respondents indicated two time-related reasons as influencing their decision to stop participating. Learners faced issues keeping up with deadlines and found that the course demanded too much time. Other reasons for attrition were less prevalent, but still reported by around 10 to 25\% of respondents. Notably, just under 20\% of respondents in a typical course (median proportion) stopped participating because they learned all they intended to learn.

% say how this was done (linear combination of correlation matrices; exluding any? if so, N=?)
To investigate the underlying factor structure of the reported reasons for disengaging, we conducted an exploratory factor analysis. The optimal number of factors was determined by the parallel analysis criterion and corresponding scree plot. The following five factors emerged: first, {\em general time issues} (course requires too much time; can't keep up with deadlines); second, {\em late start} (starting course late); third, {\em difficulty} (too advanced, exam too difficult); fourth, {\em format \& content} (didn't enjoy online format, confusing videos, materials not explained well), and fifth, {\em goals \& expectations} (only exploring, learned all I wanted, not what I'm looking for, not challenging, credential not valuable). The technical difficulties item failed to load onto any factor.

% NEED TO SAY THAT IT'S LITTLE DATA HERE (n=?)
We expected reasons for disengaging from the course to vary by learner characteristics ($RQ3b$), such as learners' motivations for enrolling in the course. A correlation matrix of learners' enrollment intentions and their subsequent reasons for disengaging revealed a small number of notable associations. Taking the course with friends or colleagues was correlated with disengaging because the course was not challenging enough, the material was too advanced for the learner's background, and the credential was not valuable enough to the learner ($r=.24, .20, .16$, respectively). Not surprisingly, learners who enrolled to experience an online course tended to stop participating because they were only exploring the course ($r=.17$). Learners looking to improve their English tended to experience technical difficulties ($r=.17$), and those who enrolled for personal growth and enrichment tended to struggle to keep up with deadlines ($r=.16$).

% other covariates: e.g. do we see geolocation difference for language and tech?
% also gender differences?

@@

\subsection{Discussion}
% need to adjust once the rest is finalized above
A number of striking findings emerged from the preceeding analyses. We found strong evidence for gender inequalities in attrition and, to some extent, performance. Women were less likely to persist, even when adding a number of other demographic, geographic, and preference indicators to the model, which highlights the robustness of the relationship. The MOOCs in our dataset were on diverse topics, including some on topics in the humanities and business with high rates of participation from women. Besides gender differences, attrition was substantially higher for learners from Africa, Asia, and South America, compared to North America, Oceania, and Europe, in terms of lectures watched. We also identified a large achievement gap in terms of grades between learners from these continents. The current generation of MOOCs is seemingly not immune to inequalities that have pervaded traditional educational settings. It remains unclear, however, to what extent these gaps result from differences in Internet access and language barriers, or potentially, from experiences of psychological threat to learners' identity or sense of social belonging \cite{walton2007question}.

Levels of self-reported satisfaction were generally high and did not vary much by level of persistence, although learners who reportedly disengaged from the course tended to be less satisfied than those who reportedly persisted until the end. Learners' reasons for disengaging from the course clustered into five key factors: general time issues; problems due to starting the course late; course difficulty; issues with the course format or content, and a mismatch or fulfillment of expectations. General time issues were overall the most pronounced. Attrition in community colleges has been interpreted as a sign of success in prior work. This view resonates in the MOOC setting, where around 20\% of learners who disengaged reportedly had achieved their goal, and where progress in the course and learner satisfaction were only weakly related.

@@

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. This study investigates reasons for attrition based on psychological measures and learners' own descriptions of the challenges they faced. We invited learners who were likely to drop out of a course to provide feedback. These learners were identified with a predictive model and invited to complete a survey. The survey included self-report measures of psychological constructs to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners. Moreover, most learners indicated in Study 1 that they disengaged due to a shortage of time. A variety of factors could influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments are presumably large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. Low volitional control could reduce learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV). We constructed the following hypotheses about the role of volition in attrition due to a shortage of time:

@@

@@

The particular MOOC under observation was an undergraduate-level course on an advanced topic in computer science. It was offered in 2014 through Coursera. There were  20,048 enrolled learners; 10,510 watched at least one video, and 5,912 attempted at least one assignment. The following subsections describe how disengaging learners were identified, which measures were included in the feedback survey, and how open responses were iteratively coded.

@@

The training set was composed of learners' interaction data (predictors) and disengagement states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (cf. \cite{halawa2014dropout}, for additional information on the prediction model).

To control the length of the prediction lag, the model was only provided with learners' activity data until they were absent for 14 consecutive days. In the training phase, if the learner disengaged but the model did not detect it within 14 days of their last activity, the case was counted as a false negative. This construction results in a model with features that all fall within the defined prediction lag period. The model was fit by logistic regression with two-fold learner-level and course-level cross-validation. Prediction performance measured by the area under the ROC curve (AUC) very high (0.931 for $train$, 0.929 for $test_1$, 0.922 for $test_2$). Two of the predictors in the fitted model are highly influential. The first is the number of days since the learner has last been active. The likelihood that the learner drops out increases substantially if the learner disengages for 14 days or more. However, the likelihood of such a learner re-engaging increases with the fraction of released videos the learner has viewed prior to the absence; this is the second most influential feature in the model. The fact that almost identical results were obtained for $test_1$ and $test_2$ suggest good transferability of the model. A learner was flagged if their dropout probability exceeded 50\%. Out of the 20,048 enrolled learners in the course, only those 10,510 were considered who watched at least one lecture video, and 6,074 of them were flagged as likely to drop out. The survey was sent out to 6,053 learners (excluding 21 learners with invalid email addresses or duplicate accounts).

@@

Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked. These items served as measures of personal success. They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of. This question was asked prior to any survey questions that could suggest particular reasons for disengaging.

Two research assistants independently developed codebooks for the resulting 448 non-empty open responses. Their codebooks were consolidated and applied on Amazon Mechanical Turk (AMT) in three iterations. The codebook was updated in each iteration to reliably fit the open response data. Specific updates to the initial codebook were informed by code frequency, code correlations, and inter-coder agreement. In the first iteration, 250 randomly selected responses were each coded by four `classification experts,' which is an AMT-specific qualification. In the second iteration, the remaining responses were coded in the same way. In the final iteration, all responses were coded by two classification experts (ties were resolved by two researchers).\footnote{For additional details on the coding procedure and codebooks, see !!!!!!![URL].} The iterative coding yielded 399 relevant coded responses, which were analyzed in combination with the remaining survey data.

% Learners also rated the extent to which their progress was hindered by a number of obstacles. 
Learners self-reported social belonging (17 items on their sense of social and academic fit \cite{walton2007question}, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the stability of intelligence and talent\footnote{Adpated from \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

@@

Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners. Success was measured by relative progress and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction, but those who were successful in terms of progress. The pattern was reversed for growth mindset, such that learners who were successful in terms of progress (but not satisfaction) exhibited a stronger growth mindset. This pattern could reflect a difference in the underlying process by which these psychological factors affect learner behavior and perception. The observed differences are expected to be larger in the full population of learners, compared to the small, homogenous sample in this study, consisting of learners unlikely to persist.

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. exhibit lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$). Reasons for learner attrition were measured using iteratively coded open responses. Table \ref{tab:s2reas} provides each coded reason and percentage of open responses that indicated it. A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$). Hence, if a learner had not enough time, this tended to be the primary challenge.

To test hypotheses $H4ab$, we examined the 336 open responses that indicated having too little time. It revealed that no particular commitment was specified in about half of them (53\%). This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to capture professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses $H4ab$: that is, while the large group of learners who reported having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.

\begin{table}[h!]
\caption{Iteratively-Coded Reasons for Attrition ($n=399$)}
\label{tab:s2reas}
\small
\center
\begin{tabular}{lr}
\toprule
What challenges have you experienced while taking this course?  & \% \\
\midrule
Not enough time for the course & 84 \\
The course format or teaching style is not a good fit & 18 \\
The level of the course is too advanced & 10 \\
The course is not interesting or valuable enough & 7 \\
Additional course materials needed to learn the topic & 4 \\
Technical difficulties & 2 \\
Language barrier & 1 \\
\bottomrule
\end{tabular}
\end{table}

@@

In this follow-up study, we examined the role of psychological factors in attrition and further investigated reasons for attrition. Personal success in a MOOC was associated with three established psychological factors: goal striving, social belonging, and growth mindset. Prior work in more traditional achievement-oriented environments found reduced levels of these psychological factors to result in lower persistence and performance. The current observational findings highlight the critical role of these factors in an online learning environment. Learners with higher levels of goal striving, social belonging, and growth mindset reported greater personal success ($H1,H2,H3$). An online learner with, for instance, a strong sense of social belonging may derive greater satisfaction from the course and persist for longer than one with a weak sense of belonging. These psychological factors tend to vary across poeple based on prior experiences and they can change as a result of contextual cues. Pschological factors could thereby, at least partially, account for individual differences in persistence and performance (Study 1).

The methodology used in this study can be extended to deliver timely and targeted interventions. The predictive model was successful in identifying a large number of learners who were facing challenges with taking the course. The majority of them were unsuccessful in their own terms. All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners. Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}). Cultivating a growth mindset was found to help community college students persist in prior work, and our findings provide correlational evidence that subjectively successful learners hold a more fluid notion of intelligence than their less successful peers. Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their social belonging.

Through iterative coding, we arrived at a list of challenges faced by a subset of learners in a particular MOOC (Table \ref{tab:s2reas}). The list maps onto the factors identified in Study 1 (general time issues, late start, difficulty, format \& content, expectations, and technical difficulties), with the additional challenge of a language barrier, which was not included as a response option in Study 1. The majority of learners indicated that they had not enough time: some struggled due to higher priority obligations that took up too much time, while others were hindered by low levels of volitional control ($H4a$). A learner needs to be highly self-regulated to be successful in a MOOC. Very few learners explicitly indicated issues with volitional control in their open responses. Learners may not possess the metacognitive skills to consciously notice this issue or they preferred not to admit to this undesirable trait. Learners who reported a shortage of time without providing specific reasons were more likely to have insufficient volitional control ($H4b$). This could be a useful proxy for identifying issues with volitional control in future work. Overall, the results suggest that learners' capacity for volitional control can be a critical ingredient for learners' success and warrants further investigation \cite{corno2001volitional}.
